---
title: "Assignment 1b: Replicating *Sexual Orientation and Labor Economics* by Cara Brown (1998)"
author: "Daniel Sánchez"
date: "Due Jan. 19th 2023"
format: pdf
knitr:
  opts_chunk: 
    echo: true
---

This document walks through the replication of Cara Brown's 1998 paper *Sexual Orientation and Labor Economics* published at *Feminist Economics*.

# Preliminaries

I need the following R libraries to make this code work:

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse) # For data wrangling
library(haven) # Reading stata files
```

I load the data from the given file below:

```{r}
#| label: data-load
df<-
  read_dta('data/Census_2016_Hierarchial.dta')
```

# Exploratory Analysis & Data Wrangling

I run some exploratory analysis which motivates my wrangling/cleaning of the data set to make it fit for the analysis below. 

```{r}
#| label: prelim-cleaning

# First, rename the income variable (empin) to something more usable

df<-
  df %>% 
  rename(income = 'empin')

# Now, replace the invalid values (88,888,888, 99,999,999) with actual NAs in R.

df <-
  df %>% 
  mutate(income_clean = 
    case_when(
      income %in% c(88888888, 99999999) ~ as.numeric(NA),
      TRUE ~ income
    )
  )

```


I now group by household and count the number of individuals in them.

```{r}
#| label: prelim-cleaning

households <-
  df %>% 
  group_by(HH_ID) %>% 
  summarise(people = n())
```






