---
title: "Assignment 4b: SUR-Demand"
author: "Daniel Sánchez Pamiño"
format: pdf
knitr:
  opts_chunk:
    warning: false
    message: false
    cache: false
---

This document walks through the results of my code. Please see the code in the `assignment4b.R` file attached to the Canvas submission, or look at the last few pages of this document for my code. 

```{r}
#| label: code
#| echo: false
#| include: false
#| cache: false

### -- Assignment 4b: Seemingly Unrelated Regressions---Demand -- ###

# Daniel Sánchez
# ECON836
# Due March 8th, 2023

# Preliminaries -------------------------------------------------------------------------------------------

# Load libraries

library(tidyverse)
library(micEconAids)
library(systemfit)
library(huxtable)
library(kableExtra)

# Load the data (from the last assignment)

load('~/GitHub/econ836/assignment4-demandestimation/data/df.RData')

# Running AIDS --------------------------------------------------------------------------------------------

# I will need to compute expenditure shares, and I do so below:

total_expenditures <-
  matched_dataset %>% 
  filter(exp_type != 'Clothing',
         expenditure >= 0) %>% 
  group_by(household) %>% 
  summarise(total_exp = sum(expenditure))

# Left join to the other dataset

df <- 
  matched_dataset %>%
  filter(exp_type != 'Clothing',
         expenditure > 0) %>% 
  left_join(total_expenditures, by = c('household')) %>% 
  mutate(exp_share = (abs(expenditure)/abs(total_exp)))

# Verify expenditure shares across households sum up to one

df %>% 
  group_by(household) %>% 
  summarise(exp = sum(exp_share))

# We will need a wide form dataset to implement this. I will also only deal with prime-aged households (the household head)
# Using only three rooms, and male household heads. Only size 2.

df_wide <-
  df %>%
  pivot_wider(names_from = exp_type,
              values_from = c(exp_share, price, expenditure)) %>% 
  left_join(total_expenditures, by = 'household') %>% 
  filter(!is.na(price_Food),
         !is.na(price_Shelter),
         !is.na(price_Transportation))

# Define the arguments necessary to run the AIDS model

priceNames <- c('price_Food', 'price_Shelter', 'price_Transportation')

shareNames <- c('exp_share_Food', 'exp_share_Shelter', 'exp_share_Transportation')

# Run the AIDS model as follows:

# aids <- aidsEst(priceNames,
                #shareNames,
                #'total_exp',
                #data = as.matrix(df_wide))

# Run a SUR model, as per AIDS

eq1 <- log(exp_share_Food) ~ log(price_Food) + log(size) + log(rooms) + as.factor(year) + as.factor(province)
eq2 <- log(exp_share_Transportation) ~ log(price_Transportation) + log(size) + log(rooms) + as.factor(year) + as.factor(province)
eq3 <- log(exp_share_Shelter) ~ log(price_Shelter) + log(size) + log(rooms) + as.factor(year) + as.factor(province)

# I include some demographic characteristics as well as year and province fixed effects

# Run the SUR model:

aids <-
  systemfit(list(eq1, eq2, eq3),
            data = df_wide,
            method = 'SUR')

# Crawford Paper ------------------------------------------------------------------------------------------

# This will require me to create new price vectors from the GST increase

# I do so below, as well as estimate the consumption of quantities by dividing by price

# THis is an estimate of the "demand"

df_wide_crawford <-
  df_wide %>% 
  mutate(
    price_Transportn = price_Transportation*1.05,
    price_Foodn = price_Food*1.05,
    price_Sheltern = price_Shelter*1.05,
    q_Transport = expenditure_Transportation/price_Transportation,
    q_Food = expenditure_Food/price_Food,
    q_Shelter = expenditure_Shelter/price_Shelter,
    qn_Transport = q_Transport*price_Transportn,
    qn_Shelter = price_Sheltern*q_Shelter,
    qn_Food = price_Foodn * q_Food
  )

# Now, we implement crawford by doing a COLI calculation as per the formula

df_wide_crawford <-
  df_wide_crawford %>%
  mutate(
    col_increase = 
      ((price_Transportn * qn_Transport + price_Foodn*qn_Food + price_Sheltern*qn_Shelter)/
         (price_Transportation * q_Transport + price_Food*q_Food + price_Shelter*q_Shelter))
  )

# Estimate the average COL and report

average_col <-
  df_wide_crawford %>% 
  summarise(MeanCol = (col_increase %>% mean(na.rm = F)) - 100)

average_col_province <-
  df_wide_crawford %>% 
  group_by(province) %>% 
  summarise(MeanCol = (col_increase %>% mean(na.rm = F)) - 100) %>% 
  mutate(province = case_when(
    province == 10 ~ 'Newfoundland and Labrador',
    province  == 13 ~ 'New Brunswick',
    province  == 24 ~ 'Quebec',
    province  == 48 ~ 'Alberta',
    province  == 11 ~ 'Prince Edward Island',
    province == 59 ~ 'British Columbia',
    province == 47 ~ 'Saskatchewan',
    province == 35 ~ 'Ontario',
    province == 12 ~ 'Nova Scotia',
    province == 46 ~ 'Manitoba'
  )
  )
```

# Method 1: An Ideal Demand System

I present the results of the AIDS estimation by seemingly unrelated regression below:

# Method 2: Crawford Paper

I estimate the COLI approach from the 2022 Crawford paper. Note that this does not imply estimating any kind of regression, so I only present the average COL increase in % terms by province across all years. Please review my code if you need to know that I ran the actual model. I followed the given formula below:

$$ 1 + \pi = \dfrac{q'\cdot p'}{q \cdot p}$$
where variables with a prime are the ones estimated after the tax increase. The overall average across the years is `r average_col`%. 

```{r}
#| label: crawford
#| results: asis
kable(average_col_province,
      format = 'latex',
      booktabs = TRUE,
      longtable = T)
```

